---
title: "main_functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{main_functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(wasted)
```

```{r}
get_country_summary(plastics_data, country, year = NULL)
```

This function returns a tibble with key stats for a given country: total plastic, plastic per capita, GDP per capita, rank among all countries, and breakdown by plastic type. year defaults to both years if NULL.

```{r}
plot_continent_map(plastics_data, continent, year = NULL, per_capita = TRUE)
```

This function should render an interactive Leaflet map of recorded plastic waste, filtered to a single
continent and zoomed to that region automatically.

```{r}
plot_gdp_vs_plastic(plastics_data, region = NULL, year = NULL, log_scale = TRUE)
```

This function renders a scatterplot of log-transformed total recorded plastic waste against GDP per
capita, colored by world region. Useful for exploring whether wealthier countries tend
to record more or less plastic waste, and how that relationship varies across continents.
