---
title: "Main Functions"
vignette: >
  %\VignetteIndexEntry{Main Functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(wasted)
data <- load_data()
```

## `compute_country_stats()`

Aggregates the raw plastic waste data to the country level, computing totals,
per capita figures, and global ranks. This is the foundation for the other
summary functions.

```{r}
compute_country_stats(data)
```

You can optionally filter to a single year:

```{r}
compute_country_stats(data, year = 2019)
```

## `get_country_summary()`

Returns a tibble with key statistics for a given country: total plastic waste,
plastic per capita, GDP per capita, global ranks, and a breakdown by plastic
type. If `year` is `NULL` both years are combined.

```{r}
get_country_summary(data, "Montenegro")
```

```{r}
get_country_summary(data, "Philippines", year = 2020)
```

## `plot_plastic_vs_metric()`

Renders a scatterplot of total recorded plastic waste against a chosen
country-level metric (`gdp_per_capita` or `population`), colored by world
region. Useful for exploring whether wealthier or more populous countries tend
to record more plastic waste, and how that relationship varies across
continents.

```{r}
plot_plastic_vs_metric(data)
```

You can filter to a specific region or year using the `year` and `region` arguments. You may also switch both the metric (default `gdp_per_capita` with `population` as an alternative option), and whether the data is displayed on the logarithmic scale using `log_scale` (defaults to TRUE).

```{r}
plot_plastic_vs_metric(data, metric = population, region = "Asia")
```

```{r}
plot_plastic_vs_metric(data, metric = population, year = 2020, log_scale = FALSE)
```

## `plastic_type_summary()`

Calculates the proportion of each plastic type out of the total recorded
plastic waste. Both years are combined by default, but can be filtered
using the optional `region` and `year` arguments.

```{r}
plastic_type_summary(data)
```

Optionally filter to a region and/or year using `region` and `year` arguments.

```{r}
plastic_type_summary(data, region = "Asia")
```

```{r}
plastic_type_summary(data, year = 2020)
```

## `plastic_by_company()`

Returns the top companies by total recorded plastic waste. Results can be
filtered by region or year, and the number of companies returned controlled
with `top_n` (defaults to 10).

```{r}
plastic_by_company(data)
```


Optionally filter to a region and/or year using `region` and `year` arguments.

```{r}
plastic_by_company(data, top_n = 5, region = "Europe")
```

```{r}
plastic_by_company(data, year = 2019)
```

## `plot_plastic_by_company()`

Creates a bar chart of the companies with the highest recorded plastic waste
totals across all countries and all years available in the tidytuesday dataset.

```{r}
plot_plastic_by_company(data)
```

## `fit_plastic_lmer()`

Fits a linear mixed effects model predicting log total plastic waste from log
GDP per capita and world region, with a random intercept for country. Returns
a formatted table of the fixed effects on region, log GDP per capita and their interaction.

```{r}
fit_plastic_lmer(data)
```

## Internal Validation Functions

`validate_year()` and `validate_region()` are internal helper functions that
are not exported. They are called automatically by the exported functions
whenever a `year` or `region` argument is supplied, handling both input
validation and filtering in one step.

- `validate_year()` checks that `year` is one of `2019` or `2020` and filters
  the data to matching rows.
- `validate_region()` checks that `region` is one of `"Africa"`, `"Asia"`,
  `"Americas"`, `"Europe"`, or `"Oceania"` and filters the data to matching
  rows.

If an invalid value is supplied, both functions will stop with an informative
error message before any computation takes place.

